---
title: "Host Preference Test"
author: "Priya Shahi"
date: "2026-04-23"
output: 
  word_document: 
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
  md_document:
    variant: gfm
  pdf_document:
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)

library(ggplot2)
library(tidyverse)
library(lme4)
library(emmeans)
library(dplyr)
library(multcompView)
library(multcomp)
```

---

# Overview

This analysis examines the host preference of *Bemisia tabaci* (sweetpotato whitefly) in 
dual-choice bioassays comparing aromatic herb companions (Mint, Sage, Basil) against Cotton 
as the primary host. Eggs and nymphs deposited on each plant type serve as proxies for 
oviposition preference and nymphal establishment, respectively.

Three separate choice arenas were tested:

- **Arena 1:** Mint vs. Cotton  
- **Arena 2:** Sage vs. Cotton  
- **Arena 3:** Basil vs. Cotton  

---

# Load and Reshape Data

```{r load-data}
raw <- read_csv("Project_data.csv")
head(raw)
```

```{r reshape}
names(raw)
find("select")

# Arena 1: Mint vs Cotton
arena1 <- raw %>%
  dplyr::select(Replication, Plant, Eggs, Nymph) %>%
  mutate(Arena = "Mint vs Cotton")

# Arena 2: Sage vs Cotton
arena2 <- raw %>%
  dplyr::select(Replication = Replication.1, Plant = Plant.1,
         Eggs = Eggs.1, Nymph = Nymph.1) %>%
  mutate(Arena = "Sage vs Cotton")

# Arena 3: Basil vs Cotton
arena3 <- raw %>%
 dplyr::select(Replication = Replication.2, Plant = Plant.2,
         Eggs = Eggs.2, Nymph = Nymph.2) %>%
  mutate(Arena = "Basil vs Cotton")

# Combine all arenas
df <- rbind(arena1, arena2, arena3) %>%
  mutate(
    Plant       = as.factor(Plant),
    Arena       = as.factor(Arena),
    Replication = as.factor(Replication)
  )

head(df)
glimpse(df)
```

---

# Summary Statistics

```{r summary-stats}
summary_stats <- df %>%
  group_by(Arena, Plant) %>%
  summarise(
    n          = n(),
    mean_eggs  = mean(Eggs,  na.rm = TRUE),
    sd_eggs    = sd(Eggs,    na.rm = TRUE),
    se_eggs    = sd_eggs / sqrt(n),
    mean_nymph = mean(Nymph, na.rm = TRUE),
    sd_nymph   = sd(Nymph,   na.rm = TRUE),
    se_nymph   = sd_nymph / sqrt(n),
    .groups    = "drop"
  )

knitr::kable(summary_stats, digits = 2,
             caption = "Summary statistics: Mean eggs and nymphs per plant type per arena")
```

---

# Preference Index (PI)

The Preference Index quantifies the relative attractiveness of each plant in a dual-choice arena:

$$PI = \frac{N_{herb} - N_{cotton}}{N_{herb} + N_{cotton}}$$

PI > 0: preference for herb; PI < 0: preference for Cotton; PI = 0: no preference.

```{r preference-index}
pi_data <- df %>%
  group_by(Arena, Replication, Plant) %>%
  summarise(Eggs = sum(Eggs), Nymph = sum(Nymph), .groups = "drop") %>%
  pivot_wider(names_from = Plant, values_from = c(Eggs, Nymph)) %>%
  mutate(
    herb_eggs = case_when(
      Arena == "Mint vs Cotton" ~ Eggs_Mint,
      Arena == "Sage vs Cotton" ~ Eggs_Sage,
      Arena == "Basil vs Cotton" ~ Eggs_Basil
    ),
    herb_nymph = case_when(
      Arena == "Mint vs Cotton" ~ Nymph_Mint,
      Arena == "Sage vs Cotton" ~ Nymph_Sage,
      Arena == "Basil vs Cotton" ~ Nymph_Basil
    )
  )


# Compute PI per arena dynamically
pi_calc <- df %>%
  group_by(Arena, Replication) %>%
  summarise(
    herb_plant  = Plant[Plant != "Cotton"][1],
    herb_eggs   = Eggs[Plant != "Cotton"],
    cotton_eggs = Eggs[Plant == "Cotton"],
    herb_nymph  = Nymph[Plant != "Cotton"],
    cotton_nymph = Nymph[Plant == "Cotton"],
    .groups = "drop"
  ) %>%
  mutate(
    PI_eggs  = (herb_eggs  - cotton_eggs)  / (herb_eggs  + cotton_eggs  + 1e-6),
    PI_nymph = (herb_nymph - cotton_nymph) / (herb_nymph + cotton_nymph + 1e-6)
  )

pi_summary <- pi_calc %>%
  group_by(Arena, herb_plant) %>%
  summarise(
    mean_PI_eggs  = mean(PI_eggs,  na.rm = TRUE),
    se_PI_eggs    = sd(PI_eggs,    na.rm = TRUE) / sqrt(n()),
    mean_PI_nymph = mean(PI_nymph, na.rm = TRUE),
    se_PI_nymph   = sd(PI_nymph,   na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

knitr::kable(pi_summary, digits = 3,
             caption = "Preference Index (PI): negative = preference for Cotton")
```

---

# Figures

## Figure 1 — Mean Eggs Deposited per Plant Type per Arena

```{r fig-eggs, fig.cap="Mean number of eggs deposited on herb vs. Cotton across three choice arenas. Error bars represent ± 1 SE."}
egg_plot <- ggplot(summary_stats,
       aes(x = Plant, y = mean_eggs, fill = Plant)) +
  geom_col(color = "black", width = 0.6) +
  geom_errorbar(aes(ymin = mean_eggs - se_eggs,
                    ymax = mean_eggs + se_eggs),
                width = 0.2, linewidth = 0.6) +
  facet_wrap(~ Arena, scales = "free_x") +
  scale_fill_manual(values = c(
    "Mint"   = "#4CAF50",
    "Sage"   = "#8BC34A",
    "Basil"  = "#2E7D32",
    "Cotton" = "#E8C96A"
  )) +
  labs(
    title    = expression(paste(italic("Bemisia tabaci"), " Oviposition Preference — Dual Choice Bioassay")),
    x        = "Plant Species",
    y        = "Mean Eggs Deposited (± SE)",
    fill     = "Plant"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title       = element_text(hjust = 0.5, face = "bold", size = 11),
    strip.background = element_rect(fill = "grey90", color = "black"),
    strip.text       = element_text(face = "bold"),
    legend.position  = "none"
  )

egg_plot

```

## Figure 2 — Mean Nymphs per Plant Type per Arena

```{r fig-nymphs, fig.cap="Mean number of nymphs recorded on herb vs. Cotton across three choice arenas. Error bars represent ± 1 SE."}
nymph_plot <- ggplot(summary_stats,
       aes(x = Plant, y = mean_nymph, fill = Plant)) +
  geom_col(color = "black", width = 0.6) +
  geom_errorbar(aes(ymin = mean_nymph - se_nymph,
                    ymax = mean_nymph + se_nymph),
                width = 0.2, linewidth = 0.6) +
  facet_wrap(~ Arena, scales = "free_x") +
  scale_fill_manual(values = c(
    "Mint"   = "#4CAF50",
    "Sage"   = "#8BC34A",
    "Basil"  = "#2E7D32",
    "Cotton" = "#E8C96A"
  )) +
  labs(
    title = expression(paste(italic("Bemisia tabaci"), " Nymphal Establishment — Dual Choice Bioassay")),
    x     = "Plant Species",
    y     = "Mean Nymphs (± SE)",
    fill  = "Plant"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title       = element_text(hjust = 0.5, face = "bold", size = 11),
    strip.background = element_rect(fill = "grey90", color = "black"),
    strip.text       = element_text(face = "bold"),
    legend.position  = "none"
  )

nymph_plot

```

## Figure 3 — Preference Index by Arena

```{r fig-pi, fig.cap="Preference Index (PI) for eggs and nymphs per arena. PI < 0 indicates preference for Cotton over the aromatic herb."}
pi_long <- pi_summary %>%
  pivot_longer(
    cols = c(mean_PI_eggs, mean_PI_nymph),
    names_to = "Metric",
    values_to = "PI"
  ) %>%
  mutate(
    se = ifelse(
      Metric == "mean_PI_eggs",
      pi_summary$se_PI_eggs[match(Arena, pi_summary$Arena)],
      pi_summary$se_PI_nymph[match(Arena, pi_summary$Arena)]
    ),
    Metric = recode(Metric,
      mean_PI_eggs = "Eggs",
      mean_PI_nymph = "Nymphs"
    )
  )
pi_plot <- ggplot(pi_long, aes(x = Arena, y = PI, fill = Metric)) +
  geom_col(position = position_dodge(0.7), width = 0.6, color = "black") +
  geom_errorbar(aes(ymin = PI - se, ymax = PI + se),
                position = position_dodge(0.7), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.7) +
  scale_fill_manual(values = c("Eggs" = "#5B8DB8", "Nymphs" = "#F4A460")) +
  labs(
    title = "Preference Index: Aromatic Herb vs Cotton",
    x     = "Choice Arena",
    y     = "Preference Index (PI)",
    fill  = "Life Stage"
  ) +
  coord_flip() +   # ✅ THIS MAKES IT HORIZONTAL
  theme_classic(base_size = 12) +
  theme(
    plot.title      = element_text(hjust = 0.5, face = "bold"),
    axis.text.x     = element_text(angle = 0),  # no need to tilt anymore
    legend.position = "right"
  )

pi_plot

```

---

# Statistical Analysis

## Wilcoxon Signed-Rank Test



```{r Wilcox-eggs}
# ---- EGGS ----
 cat("=== ARENA 1: Mint vs Cotton (Eggs) ===\n")

mint_eggs <- df %>%
  filter(Arena == "Mint vs Cotton") %>%
  dplyr::select(Replication, Plant, Eggs) %>%
  pivot_wider(names_from = Plant, values_from = Eggs)

print(wilcox.test(mint_eggs$Mint, mint_eggs$Cotton, paired = TRUE))


cat("\n=== ARENA 2: Sage vs Cotton (Eggs) ===\n")

sage_eggs <- df %>%
  filter(Arena == "Sage vs Cotton") %>%
  dplyr::select(Replication, Plant, Eggs) %>%
  pivot_wider(names_from = Plant, values_from = Eggs)

print(wilcox.test(sage_eggs$Sage, sage_eggs$Cotton, paired = TRUE))


cat("\n=== ARENA 3: Basil vs Cotton (Eggs) ===\n")

basil_eggs <- df %>%
  filter(Arena == "Basil vs Cotton") %>%
  dplyr::select(Replication, Plant, Eggs) %>%
  pivot_wider(names_from = Plant, values_from = Eggs)

print(wilcox.test(basil_eggs$Basil, basil_eggs$Cotton, paired = TRUE))
```

```{r Wilcox-nymphs}
# ---- NYMPHS ----
cat("\n=== ARENA 1: Mint vs Cotton (Nymph) ===\n")

mint_nymph <- df %>%
  filter(Arena == "Mint vs Cotton") %>%
  dplyr::select(Replication, Plant, Nymph) %>%
  pivot_wider(names_from = Plant, values_from = Nymph)

print(wilcox.test(mint_nymph$Mint, mint_nymph$Cotton, paired = TRUE))


cat("\n=== ARENA 2: Sage vs Cotton (Nymph) ===\n")

sage_nymph <- df %>%
  filter(Arena == "Sage vs Cotton") %>%
  dplyr::select(Replication, Plant, Nymph) %>%
  pivot_wider(names_from = Plant, values_from = Nymph)

print(wilcox.test(sage_nymph$Sage, sage_nymph$Cotton, paired = TRUE))


cat("\n=== ARENA 3: Basil vs Cotton (Nymph) ===\n")

basil_nymph <- df %>%
  filter(Arena == "Basil vs Cotton") %>%
 dplyr::select(Replication, Plant, Nymph) %>%
  pivot_wider(names_from = Plant, values_from = Nymph)

print(wilcox.test(basil_nymph$Basil, basil_nymph$Cotton, paired = TRUE))
```

